Music Recommendation System Using the Million Song Dataset

### 2015-05-08 writeup/MSDgroupwriteup.tex
Updating .tex

\subsection{Probabilistic Matrix Factorization}
Matrix factorization is a type of collaborative filtering algorithm that can be used to build a recommendation system for users based on user feedback of objects (in this case songs). This allows us to recommend songs to users based on their listening history without the need for content based approaches. 
Training and testing matrices were constructed, of which both have $N_1$ users (rows denoted by $u_i$) and $N_2$ songs (columns denoted by $v_j$). As users have only listened to a fraction of all the songs available, the matrices are sparse, containing zeros in all entries except for those in which a user (representing a single row) has listened to. The goal of matrix factorization is to factor the training matrix into the product of two matrices $U$ and $V$.  The matrix $U$ is $N_1 \times d$ and the matrix $V$ is $d \times N_2$. Another goal of matrix factorization is typically to learn a low-rank factorization of the original matrix while restricting the patterns that are generated in the product of the factorized matrices (e.g. if a user likes a song in a specific group of songs, the user may also like other songs in that group of songs). The choice of $d$ can be subjective but $20$ is a common value to start with. In the factorized matrices, the predicted rating will be $\hat{M}_{ij} = u_i^\TRANSPOSE  v_j$
Using a coordinate ascent algorithm over 100 iterations, each row ($u_i$) and column ($v_j$) of the training matrix is then updated (equations \ref{eq1} and \ref{eq2}) in order to maximize the log joint likelihood (equation \ref{eq3}).

@@

There are several hyperparameters that can be tweaked in order to achieve a better MAP result using matrix factorization. The main ones that we explored were the variance, the rank $d$ of the factorization, and the number of iterations. As the project was being done using Python and the main computational package (\emph{scikit-learn}) did not have probabilistic matrix factorization functions, we decided to apply the algorithm that we wrote ourselves for a different class \citep{koren2009matrix}.

>>>>

\subsection{Probabilistic Matrix Factorization}
Probabilistic matrix factorization (PMF) is a type of collaborative filtering algorithm that can be used to build a recommendation system for users based on user feedback of objects (in this case songs). This allows us to recommend songs to users based on their listening history without the need for content based approaches. 
Training and testing matrices were constructed, of which both have $N_1$ users (rows denoted by $u_i$) and $N_2$ songs (columns denoted by $v_j$). As users have only listened to a fraction of all the songs available, the matrices are sparse, containing zeros in all entries except for those in which a user (representing a single row) has listened to. The goal of matrix factorization is to factor the training matrix into the product of two matrices $U$ and $V$.  The matrix $U$ is $N_1 \times d$ and the matrix $V$ is $d \times N_2$. Another goal of matrix factorization is typically to learn a low-rank factorization of the original matrix while restricting the patterns that are generated in the product of the factorized matrices (e.g. if a user likes a song in a specific group of songs, the user may also like other songs in that group of songs). The choice of $d$ can be subjective but $20$ is a common value to start with. In the factorized matrices, the predicted rating will be $\hat{M}_{ij} = u_i^\TRANSPOSE  v_j$
Using a coordinate ascent algorithm over 100 iterations, each row ($u_i$) and column ($v_j$) of the training matrix is then updated (equations \ref{eq1} and \ref{eq2}) in order to maximize the log joint likelihood (equation \ref{eq3}).

@@

There are several hyperparameters that can be tweaked in order to achieve a better MAP result using matrix factorization. The main ones that we explored were the variance, the rank $d$ of the factorization, and the number of iterations. As the project was being done using Python and the main computational package (\emph{scikit-learn}) did not have probabilistic matrix factorization functions, we decided to apply the algorithm that we wrote ourselves for a different class \citep{koren2009matrix}. It should be noted that while PMF algorithms can be parallelized to significantly improve running time and scaled to handle large amounts of data, being limited to a single machine to run these computations meant that sub setting the data was necessary in order to actually execute the algorithm successfully while limiting computational time to durations that were feasible for exploration.
